Adaptative Computation:\\ An Essay in Memory of John H. Holland

### 2015-11-03 paper.tex
second try on commit

%\title{Adaptation Writ Large:\\ An Essay in Memory of JohnH. Holland}
\title{Adaptative Computation:\\ An Essay in Memory of John H. Holland}

@@

John was a direct intellectual descendant of the cybernetics era, and
early on was strongly influenced by the work of von Neumann, Wiener,
Ashby, and Turing, all of whom viewed computation as a broad,
interdisciplinary enterprise.  Thus, John became an early proponent of
interdisciplinary approaches to computer science and was an active
evangelist of what is now called {\emph computational thinking}, reaching out
enthusiastically to psychologists, economists, physicists, linguists,
philosophers, and pretty much anyone he came in contact with.  As a
result, even though he received what was arguably the world's first
computer science Ph.D. in 1959, his contributions are sometimes better known
outside of CS than within.

John Holland is best known for his invention of {\emph genetic
  algorithms} (GAs), a family of search and learning methods inspired
by biological evolution.  Since their invention in the 1960s, GAs,
along with related evolutionary computation methods, have become a
thriving subfield of computer science, with widespread scientific and
commercial applications.  While the mechanisms of GAs are well-known
to much of the CS community, fewer are aware that GAs were only one
offshoot of Holland's much broader motivation---to develop a general
theory of adaptation in complex systems.  This motivation was the
driving force in all of Holland's research.

In this short essay, we sketch five key, recurrent themes of Holland's
work on adaptive systems: (1) the evolutionary dynamics of ``building
blocks''; (2) learning via credit assignment and rule discovery; (3)
the emergence of internal models; (4) the art of scientific modeling;
and (5) the universal properties of complex adaptive systems.  We
discuss the role these themes have played in computer science, and
highlight especially the ideas that we think remain relevant to
today's research agendas.

\section{Evolutionary Dynamics of ``Building Blocks''}

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic work
on integrating genetics with Darwinian selection \cite{Fisher1930}.
As Holland further read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain changing environments and use it in
order to improve their performance with respect to those environments.
Moreover, Holland recognized that in systems with {\em adaptive
  agents} of this kind, there is never a state of {\emph equilibrium}
or a final \emph{optimum} configuration.  Due to the environment's
perpetual novelty,
%SF thinks Holland did not invent this term (as Holland termed it), 
adaptation is a continual 
open-ended process.  Holland's focus on open-ended,
non-equilibrium dynamics was in stark contrast with the mainstream
approach (at the time) in all these fields---the belief that ``solving
for'' stable equilibrium dynamics was the scientific goal.  Holland's
contrary view was that a system in stable equilibrium is essentially
\emph{dead}.

@@

Underlying Holland's theory of adaptation are three core ideas:
Population-based search, building blocks and recombination, and the
exploration-exploitation tradeoff.

\begin{itemize}
\item{\bf Populations and implicit parallelism:} A population of
  individuals can be thought of as a set of samples taken from a much larger
  search space.  Importantly, the populations are competitive, and
  the evolutionary dynamic biases the set of samples
  over time towards high-fitness regions of the search space.

@@

\item{\bf Building Blocks and recombination:} High-fitness individuals
  are discovered in stages, first by finding high-fitness partial
  solutions (the building blocks) through random sampling, and over
  time assembling the partial solutions into more complete solutions
  with even higher fitness.  For example, when searching the space of
  all $n$-bit binary strings, almost any random sample would contain
  examples of strings whose first bit is set to 1, and also many
  samples of strings whose second bit is set to 1, but there would be
  many fewer samples that have both of the first two bits set to
  1.  More compelling examples include the \emph{Krebs cycle}, a core
  cellular metabolic pathway, a group of neurons that responds to an
  oriented edge in vision, or the ``swap'' operator in a sorting
  algorithm.

 To succeed, such a search requires first that
  high-fitness subparts be preserved once found, and second, that
  there be an efficient mechanism for combining promising building
  blocks into higher-order building blocks.  Holland proposed using an
  analog of Darwinian selection for the first and the idea of crossing
  over in genetics for the second.  Holland referred to the building
  blocks as \emph{schemas,} which correspond mathematically to hyperplanes in the
  search space.

%In a population undergoing adaptation,
%  individuals can be decomposed into {\emph building blocks}---sets of
% traits that are the evolutionary ``atoms'' of an individual's
%  fitness or performance.

\item{\bf Exploitation Versus Exploration:} Successful adaptation
  requires maintaining a balance between {\emph exploitation}, in which
  tried-and-true building blocks propagate in a population, and {\emph
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  
%A system that focuses too much on exploitation
%  risks never finding better individuals or not being able to adapt to
%  changing environments.  Conversely, too much focus on exploration
%  means that the system won't leverage the successful building blocks
%  that have already been discovered.  %There is some optimal balance
%  between these two modes of processing.
\end{itemize}

Several of Holland's early papers (e.g.,
\cite{Holland1962,Holland1973}) and his influential
1975 book {\emph Adaptation in Natural and Artificial Systems}
\cite{Holland1975} developed a general, formal setting in which these ideas
could be expressed mathematically.

Inspired by Bellman \cite{Bellman1961} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme {\emph
  exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme {\emph
  exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?
In \cite{Holland1973,Holland1975} Holland derived
an equation for such a strategy and argued that the optimal strategy
allocated trials to the observed best arm at a slightly higher rate than
exponential.  
% SF used 'argue' because some claim that his derivation was wrong.
He then extended this 
% Let $A$ denote the arm currently
%observed to have the higher payoff probability, and $B$ denote the
%other arm.  Holland showed that the optimal strategy (the one that
%yields maximal payoff, or minimal loss) is for the number of trials
%allocated to $A$ to grow slightly faster than an exponential function
%of the number of trials allocated to $B$.  Holland then showed that this
result to the {\emph multi-armed bandit} case.

In a population undergoing adaptation, building blocks consitute the arms of a multi-armed bandit.  Each
building block can be said to have a probability of ``payoff'' (i.e.,
contribution to a given individual's fitness).  Evaluating an
individual in an environment is like ``pulling the arms'' on a
multi-armed bandit, where the ``arms'' correspond to each of the
building blocks making up that individual.  Thus, in the process of
assigning a fitness to each individual in a population, adaptive
evolution can be seen as implicitly sampling the many building blocks
making up that collection of individuals.  The average fitness over
many individuals containing a particular building block gives an
estimate of that building block's payoff probability.

% The question of how to balance exploitation and exploration---how to
% optimally allocate trials to different arms based on their observed
% payoff---now becomes the question of how to optimally sample in the
% vast space of possible building blocks, based on their observed
% average fitness.  Of course adaptive evolution deals in populations of
% individuals, not building blocks.  There is no explicit mechanism for
% keeping statistics on building blocks' ``observed average fitness''.
% However, Holland's central idea here is that an approximation to
% optimal building-block sampling does indeed occur, as an emergent
% property of the population dynamics.

Holland argued that an approximation to
optimal building-block sampling actually occurs, as an emergent
property of the population dynamics defined by a
% To show this mathematically, he defined an idealization of
%adaptive evolution: an algorithm he called a {\emph reproductive
% plan}.
a population-based stochastic process
operating on bit strings, involving the ``genetic'' operators of
 reproduction, fitness-based selection, crossover, mutation, and
 inversion.  
%Building blocks are formalized as {\emph
%   schemata}---patterns of bits within a string. (Here are two examples
% of schemas: ``all strings beginning with the bits 10'' or ``all
% strings that start with the pattern 1*1'', where * can be replaced by
% either 0 or 1).

% Holland's most important result was the following: He proved
% mathematically that his reproductive plan results---implicitly---in a
% near-optimal allocation of trials to schemata, thus optimizing the
% exploitation versus exploration balance.  Holland noted that the
% explicit act of assigning fitnesses to the individuals in a population
% was actually implicitly sampling a much larger collection of building
% blocks.  Holland termed this {\emph implicit parallelism}; the
% parallel leveraging of statistics from implicit sampling is a main
% strength of his population-based approaches to search.

It was this formalization
% and resulting theorems that led to the
that led to the invention of genetic algorithms, which featured stochastic
population-based search and recombination as a critical
%{\emph crossover} as a critical
operation that allowed successful building blocks to be recombined and
tested in new contexts.  What made GAs unique
among other evolution-inspired algorithms at the time were the
mathematical foundations described above, the emphasis on populartions
and recombination as a central mechanism,
and a focus (at least in Holland's mind) on continual adaptation to
non-stationary environments rather than optimization to static
environments.

@@

\section{Internal Models and Perpetual Novelty}

@@

Internal models are central to Holland's theory of adaptive
systems.  He posits that all adaptive systems create and use internal
models to prosper in continually
changing environments---\emph{perpetual novelty}. Models can be tacit and learned
  through evolutionary time, as in the case of bacteria swimming up a
  chemical gradient, or explicit and learned over a single lifespan,
  as in the case of cognitive systems that incorporate experience into
  internal representations through learning.  In
  Holland's view, the key activity of an adaptive agent involves
  building and refining these data-driven models
  of the environment.
%using them to make predictions and take actions,
%  and occasionally receiving positive or negative feedbacks
%  (intermittent rewards).

In his second book, \emph{Induction} \cite{Holland1989}, Holland and
his co-authors tackled the question of how cognitive agents can
possibly learn and profit from
internal models, without supervision by combining
environmental inputs and rewards with stored knowledge.   In their
framework, a 
model defines a set of equivalence relations over environmental states,
together with a set of transition rules, which are learned over time
based on environmental rewards (or punishments).  Models that
form valid homomorphisms with the environment allow the system to make
accurate predictions.  In Holland's conception, the equivalence
classes are initially very general, say ``moving object'' and
``stationary object,'' and over time, through experience and learning,
 are specialized into more useful and precise subclasses, say
``insect'' and ``nest.''   Over time, the adaptive system
builds up a default hierarchy of rules covering general
cases and refinements for specific classes.

@@

% Maybe this should be moved to section on Complexity
Although the idea of default hierarchies was prevalent in 
knowledge representation systems of the era, Holland made two key
contributions.  The first was his emphasis on homomorphisms as a formal way to
evaluate model validity, and  idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
% %Today, even though computational modeling is a major
% %application area of computing, essential to most natural sciences and
% %engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of to characterize when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update the model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1977,Holland1986}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning, including unsupervised learning, non-Markovian learning, AND
ONE OTHER?.

@@

% In the 1970s and 80s Holland formulated and extensively explored a
% particular model of decision-making, action, and inductive learning,
% called the {\emph classifier system} (e.g.,)
% .  In a classifier system, a population
% of ``if-then'' rules interacts with an environment, and over time is
% able to learn to improve its performance via both {\emph
%   credit-assignment} and {\emph rule-discovery} algorithms.

% Like genetic algorithms, classifier systems can be viewed both as 
% models of adaptation and as artificial-intelligence (AI) methods.  At
% the time Holland was developing classifier systems, the field of AI
% was focused on expert systems, which typically did not learn on their
% own, a fundamental deficiency in Holland's view: ``[Expert] systems
% are brittle in the sense that they respond appropriately only in
% narrow domains, requiring substantial human intervention to compensate
% for even slight shifts in domain'' \cite{Holland1986}. In
% contrast, Holland proposed that ``{\emph induction} is the basic, and
% perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1961,Samuel1959}).  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Waldrop1993} (p. 182).  
The if-then rules, when activated, compete to post their results on a
shared ``message list,'' modeling the system's short-term memory
(again inspired by Hebb's work and AI blackboard systems of the day).
Unlike the AI systems, however new rules were generated automatically,
in a trial-and-error fashion using a
genetic algorithm.

Successful rules were
strengthened over time if their predictions led to positive rewards
from the environment 
(and weakened otherwise).  
through a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules transferred some
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.   [ONE SENTENCE HERE
ABOUT SAMUEL'S CHECKER PLAYING PROGRAM.]
%, e.g., finding a target, or avoiding a predator,

%Finally, new rules for the population are discovered by a genetic
%algorithm, using the strength of rules as a fitness function.
%Classifier systems are fairly complex architectures, with learning
%occurring at multiple time scales.  
Although Holland proposed classifier systems as an executable theory
of inductive processes in cognition, other researchers took the
computational system more seriously, applying it to areas as diverse
as poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock market
\cite{Palmer1994}.  See Ref.~\cite{Booker1989} for more details about
practical applications of classifier systems.  Today, other
reinforcement learning methods are more popular for real-world
decision and control problems, but classifier systems can perhaps be
thought of as an essential ``stage-setting'' method that enabled the
development of later approaches.

@@

Holland was primarily interested in how the two
learning mechanisms (discovery of new rules and apportioning credit to
existing rules) could work together to create 
%subsets of rules
%that accurately model the environment, and thus are able to ``fire''
%at appropriate times.  
%Holland envisioned that, with a complex environment and enough time
%for learning, a classifier system might learn an intricate default
useful default hierarchies of rules.  He emphasized that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it had learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
%Offspring rules, which begin life weaker than do their parents, can
%win the competition and influence the system's behavior only when
%there are not strong rules whose conditions are satisfied---in other
%words when the system does not know what to do.  If their actions
%help, they survive; if not they are soon replaced.  
%Thus,
... new rules do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{Holland1992}.

@@

Given that Holland believed that the ability to learn and manipulate
internal models was essential for any adapting system, it is no
surprise that he viewed modeling as essential for scientific inquiry.

@@

Today, we use computational models both for \emph{prediction}---by
analyzing data via statistical models---and for \emph{understanding}
how systems work---by probing the effects of hypothesized underlying
mechanisms.  This latter use of models was dear to Holland's heart.
In his view, the key to science was understanding the mechanisms that
cause a system to behave in a certain way, an aspiration that goes
well beyond data fitting methods, which typically focus only on
describing the aggregate behavior of a system.

For example, a statistical model that describes the boom and bust
pattern of the stock market cannot address the underlying mechanisms
that lead to these cycles, through the collective actions of myriad
individual buy/sell decisions.  Similarly, the genetic algorithms for
which Holland is so famous provide a simple computational framework in
which to explore the dynamics of Darwinian evolution and whether the
basic mechanisms of variation, differential reproduction, and heredity
are sufficient to account for the richness of our natural world.

The emphasis on exploratory models to build intuitions was an
important theme of Holland's work, and he often quoted Eddington's
remark on the occasion of the first experimental test of Einstein's
theory of relativity: "The contemplation in natural science of a wider
domain than the actual leads to a far better understanding of the
actual" \cite{Eddington1927}.

Holland was interested in models that explored basic principles and
mechanisms, even if they didn't make specific or detailed predictions.
Such models can show generically how certain behaviors could be produced.
Holland pioneered a style of modeling that has come to be known as
'individual based' or 'agent based,' in which every component of a
system is represented explicitly and has state, e.g., every trader in
a stock market system or every cell in an immune system model.  In
such models, each agent has its own behavior rules, which it can
update (or learn) over time.  In order to capture the constraints of
systems living under spatial constraints, these models are often
defined over spatial structures, such as networks or simple grids.

A given agent-based model encodes a theory about the mechanisms that
are relevant for producing the behavior of interest.  Similar to
expert systems, such models are especially useful for studying systems
that don't have analytic mathematical descriptions.  Agent-based
models can facilitate interdisciplinary collaborations because the
underlying rules can be easily communicated.  The agent-based models
championed by Holland were typically idealized versions of complex
systems and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

It should be noted that Holland's view of modeling is by no means
typical.  For example, in a textbook on computational modeling, the
authors offer the following definition: “Modeling is the application
of methods to analyze complex, real-world problems in order to make
predictions about what might happen with various actions”
\cite{Shiflet2014}.   This sort of perspective completely rules out the
kind of exploratory modeling that Holland was most interested in.

>>>>

%\title{Adaptation Writ Large:\\ An Essay in Memory of John H. Holland}

\title{Adaptative Computation:\\ An Essay in Memory of John H. Holland}

@@

As a descendant of the cybernetics era, Holland was strongly influenced
by the work of von Neumann, Wiener, Ashby, and Turing, all of whom
viewed computation as a broad, interdisciplinary enterprise.  Thus,
Holland became an early proponent of interdisciplinary approaches to
computer science and an active evangelist of what is now called
\emph{computational thinking}, reaching out enthusiastically to
psychologists, economists, physicists, linguists, philosophers, and
pretty much anyone he came in contact with.  As a result, even though
he received what was arguably the world's first computer science
Ph.D. in 1959, his contributions are sometimes better known outside of
computer science than within.

Holland is best known for his invention of \emph{genetic
  algorithms} (GAs), a family of search and optimization methods inspired
by biological evolution.  Since their invention in the 1960s, GAs have
inspired many related methods and led to the thriving subfield of
\emph{evolutionary computation}, with widespread scientific and
commercial applications.  While the mechanisms of GAs are well-known,
%to much of the computer science community, 
the ideas that motivated Holland are less widely appreciated---to develop a general
%Holland's motivation 
%offshoot of Holland's much broader motivation--
theory of adaptation in complex systems.  
%This motivation was the
%driving force in all of Holland's research.

In this short essay, we consider this larger framework, sketching the
recurring themes that were central to Holland's theory of
adaptive systems: (1) discovery and 
%evolutionary 
dynamics of building blocks; (2) internal models and [prediction?] perpetual
novelty; (3) the art [SF doesn't like the term ART] of scientific modeling; and (4) common
properties of complex adaptive systems.  
%We discuss the role these
%themes have played in computer science, and highlight especially the
%ideas that we think remain relevant to today's research agendas.  [MM
%says: Are we going to do this?]

%\section{Evolutionary Dynamics of ``Building Blocks''}  
\section{Discovery and Dynamics of Building Blocks}

Holland's interest in a general theory of adaptation grew out of
his early work developing computer models of Hebbian learning
\cite{Rochester1956} and 
his reading of 
Ronald Fisher's classic work that
integrated genetics with Darwinian selection \cite{Fisher1930}.
%As Holland further read extensively in 
Over time, he recognized that adaptation was central to 
evolutionary biology,
economics, psychology, game theory, control theory, and many other fields.
%he came to recognize that
%adaptation was the most central concept in all these fields.  That is,
Each field concerns populations of agents that must continually
obtain information from uncertain changing environments and use it to
improve their performance and enhance survival.
%with respect to those environments.
Moreover, Holland recognized that adaptation must be a continual
open-ended process due to the
environment's \emph{perpetual novelty}.
%SF thinks Holland did not invent this term (as Holland termed it), 
Thus, adaptive systems never achieve a state of equilibrium or a final
optimum configuration.  This emphasis on open-ended, non-equilibrium
dynamics was in stark contrast with the mainstream approach (at the
time) in all these fields---the belief that solving for stable
equilibrium dynamics was the scientific goal.  Holland's contrary view
was that a system in stable equilibrium is essentially \emph{dead}.

@@

Underlying Holland's theory of adaptation are the following core ideas:

\begin{itemize}
\item{\bf Populations and implicit parallelism:} Adaptation occurs in
  populations of individuals evolving (or learning) over time, in
  which statistics can be leveraged to direct population dynamics.
  Evolution can be thought of as a search for populations that are highly fit with respect to their environment.  
  A population can be thought of as a parallel sampling of many
  individuals (from the space of possible individuals).  Moreover,
  that same population can be thought of as implicitly sampling a much
  space of \emph{traits} that comprise those individuals.
  Holland termed this implicit large-scale sampling of traits ``implicit
  parallelism.''    Evolutionary dynamics biases these samples over time towards
  high-fitness regions of the search space.

@@

\item{\bf Building blocks and recombination:} 
% Melanie rewrote this some to take out references to ``solutions''
  In a population undergoing adaptation,
  individuals can be decomposed into \emph{building blocks}---sets of
  traits that are the evolutionary ``atoms'' of an individual's
  fitness or performance. (As an example from biology, Holland gives
  the \emph{Krebs cycle}, a core cellular metabolic
  pathway.)   Successful individuals
  are discovered in stages, first by finding useful
  building blocks through random sampling, and over
  time recombining such building blocks to create higher-fitness individuals.

%  For example, when searching the space of
%  all $n$-bit binary strings, almost any random sample would contain
%  examples of strings whose first bit is set to 1, and also many
%  samples of strings whose second bit is set to 1, but there would be
%  many fewer samples that have both of the first two bits set to
%  1.  More compelling examples include the \emph{Krebs cycle}, a core
%  cellular metabolic pathway,

  To succeed, such a search requires first that
  useful building blocks be preserved once discovered, and second that
  there be an efficient mechanism for combining promising building
  blocks into higher-order building blocks.  Holland proposed using an
  analog of Darwinian selection for the first and the idea of crossing
  over in genetics for the second.  Holland referred to the building
  blocks as \emph{schemas.} 
%which correspond mathematically to hyperplanes in the search space. MM: this makes sense only in bit-string example.

%In a population undergoing adaptation,
%  individuals can be decomposed into \emph{building blocks}---sets of
% traits that are the evolutionary ``atoms'' of an individual's
%  fitness or performance.

\item{\bf Exploitation Versus Exploration:} Successful adaptation
  requires maintaining a balance between \emph{exploitation}, in which
 successful building blocks propagate in a population, and \emph{
    exploration}, in which existing building blocks are recombined or
  mutated in new ways.  Inspired by Bellman \cite{Bellman1961} and
others, Holland formalized this exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?   In \cite{Holland1973,Holland1975}
Holland 
%derived
%an equation for such a strategy and 
argued that the optimal strategy
allocates trials to the observed best arm at a slightly higher rate than
exponential.
% and extended the result to multi-armed bandits.
%A system that focuses too much on exploitation
%  risks never finding better individuals or not being able to adapt to
%  changing environments.  Conversely, too much focus on exploration
%  means that the system won't leverage the successful building blocks
%  that have already been discovered.  %There is some optimal balance
%  between these two modes of processing.
Next, he showed that 
populations undergoing adaptation resemble multi-armed bandits, where
each building block is an 
%undergoing adaptation constitute the 
arms in a multi-armed bandit problem.
%Each building block can be said to have a probability of ``payoff''
%(i.e., contribution to a given individual's fitness).
Thus, in the process of assigning a fitness to each individual in a
population, adaptive evolution can be seen as implicitly sampling the
many building blocks making up that collection of individuals.  The
average fitness over many individuals containing a particular building
block gives an estimate of that building block's payoff probability.

\end{itemize}

Several of Holland's early papers (e.g.,
\cite{Holland1962,Holland1973}) and his influential
1975 book \emph{Adaptation in Natural and Artificial Systems}
\cite{Holland1975} developed a general, formal setting in which these ideas
could be expressed mathematically.

It was this formalization
% and resulting theorems that led to the
that led to the invention of genetic algorithms, which featured stochastic
population-based search, as well as
%recombination as a critical
crossover between individuals as a critical
operation that allowed successful building blocks to be recombined and
tested in new contexts.  What made GAs unique
among other evolution-inspired algorithms at the time were the
mathematical foundations described above, the emphasis on populations
and recombination as central mechanisms,
and a focus 
%(at least in Holland's mind) [MM SAYS: THIS SOUNDS TOO NEGATIVE]
on continual adaptation to
non-stationary environments rather than optimization to static
environments.

@@

%\section{Internal Models and Perpetual Novelty}
\section{Internal Models and Prediction}

@@

Internal models are central to Holland's theory of adaptive
systems.  He posits that all adaptive systems create and use internal
models to prosper in the face of ``perpetual novelty'' in continually changing environments.   Models can be tacit and learned
  through evolutionary time, as in the case of bacteria swimming up a
  chemical gradient, or explicit and learned over a single lifespan,
  as in the case of cognitive systems that incorporate experience into
  internal representations through learning.  In
  Holland's view, the key activity of an adaptive agent involves
  building and refining these data-driven models
  of the environment.
%using them to make predictions and take actions,
%  and occasionally receiving positive or negative feedbacks
%  (intermittent rewards).

In his second book, \emph{Induction} \cite{Holland1989}, Holland and
his co-authors tackled the question of how cognitive agents can
possibly learn and profit from internal models by combining
environmental inputs and rewards with stored knowledge.  In their
framework, a model defines a set of equivalence relations over
environmental states, together with a set of transition rules, which
are learned over time based on environmental rewards (or punishments).
Models that form valid homomorphisms with the environment allow the
system to make accurate predictions.  In Holland's conception, the
equivalence classes are initially very general, say ``moving object''
and ``stationary object.'' Through experience and
learning, these can be specialized into more useful and precise subclasses, say
``insect'' and ``nest.''  Over time, the adaptive system builds up a
default hierarchy of rules covering general cases and refinements for
specific classes.

@@

% Maybe this should be moved to section on Complexity
Although the idea of default hierarchies was prevalent in 
knowledge representation systems of the era, Holland made two key
contributions.  The first was his emphasis on homomorphisms as a formal way to
evaluate model validity, an idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
% %Today, even though computational modeling is a major
% %application area of computing, essential to most natural sciences and
% %engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of to characterize when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update the model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1977,Holland1986}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
bucket-brigade algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning.

%including unsupervised learning, non-Markovian learning, AND
%ONE OTHER?.

@@

% In the 1970s and 80s Holland formulated and extensively explored a
% particular model of decision-making, action, and inductive learning,
% called the \emph{classifier system} (e.g.,)
% .  In a classifier system, a population
% of ``if-then'' rules interacts with an environment, and over time is
% able to learn to improve its performance via both \emph{
%   credit-assignment} and \emph{rule-discovery} algorithms.

% Like genetic algorithms, classifier systems can be viewed both as 
% models of adaptation and as artificial-intelligence (AI) methods.  At
% the time Holland was developing classifier systems, the field of AI
% was focused on expert systems, which typically did not learn on their
% own, a fundamental deficiency in Holland's view: ``[Expert] systems
% are brittle in the sense that they respond appropriately only in
% narrow domains, requiring substantial human intervention to compensate
% for even slight shifts in domain'' \cite{Holland1986}. In
% contrast, Holland proposed that ``\emph{induction} is the basic, and
% perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1961,Samuel1959}).  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebb's view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Waldrop1993} (p. 182).  
The if-then rules, when activated, compete to post their results on a
shared ``message list,'' modeling the system's short-term memory
(again inspired by Hebb's work and AI blackboard systems of the day).
Unlike the AI systems, however, new rules were generated automatically
%in a trial-and-error fashion
using a genetic algorithm.

Successful rules were strengthened over time if their predictions led
to positive rewards from the environment (and weakened otherwise)
through a credit-assignment method called the \emph{bucket-brigade}
algorithm, in which rules gaining rewards from the environment or from
other rules transferred some of their gains to those earlier-firing
``stage-setting'' rules that set up the conditions for the eventual
reward. Holland credited Arthur Samuel's pioneering work on machine
learning applied to checkers \cite{Samuel1959} as a key inspiration for these ideas.

%, e.g., finding a target, or avoiding a predator,

%Finally, new rules for the population are discovered by a genetic
%algorithm, using the strength of rules as a fitness function.
%Classifier systems are fairly complex architectures, with learning
%occurring at multiple time scales.  
Although Holland proposed classifier systems as 
an executable theory of inductive processes in cognition, other researchers took it further, 
%other researchers took the
%computational system more seriously
applying it to areas as diverse
as poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock market
\cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for more details about
practical applications of classifier systems.)  Today, other
reinforcement learning methods are more popular for real-world
decision and control problems, but classifier systems can perhaps be
thought of as an essential stage-setting method that enabled the
development of later approaches.

@@

Holland was primarily interested in how the two
learning mechanisms (discovery of new rules and apportioning credit to
existing rules) could work together to create 
%subsets of rules
%that accurately model the environment, and thus are able to ``fire''
%at appropriate times.  
%Holland envisioned that, with a complex environment and enough time
%for learning, a classifier system might learn an intricate default
useful default hierarchies of rules.  He emphasized that the
competition inherent in the learning and action mechanisms would allow
the system to adapt to a continually changing environment without
losing what it had learned in the past.  Holland put it this way:
``Competition among rules provides the system with a graceful way of
handling perpetual novelty.  When a system has strong rules that
respond to a particular situation, that is the equivalent of saying
that it has certain well-validated hypotheses.... New rules do not
interfere with the system's action in well-practiced situations but
wait gracefully in the wings as hypotheses about what to do under
novel circumstances.'' \cite{Holland1992}.

%Offspring rules, which begin life weaker than do their parents, can
%win the competition and influence the system's behavior only when
%there are not strong rules whose conditions are satisfied---in other
%words when the system does not know what to do.  If their actions
%help, they survive; if not they are soon replaced.  
%Thus,

@@

Given that Holland believed that the ability to learn and manipulate
internal models was essential for any adaptive system, it is no
surprise that he viewed modeling as essential for scientific inquiry.

@@

Today, we use computational models both for \emph{prediction}---by
analyzing data via statistical models---and for \emph{understanding}
how systems work---by probing the effects of hypothesized underlying
mechanisms.  This latter use of models was dear to Holland's heart.
In his view, the key to science was understanding the mechanisms that
cause a system to behave in a certain way, an aspiration that goes
well beyond data fitting methods, which typically focus only on
the aggregate behavior of a system.

For example, a purely statistical model that describes the boom and bust
pattern of the stock market
%cannot
does not address the underlying mechanisms
that lead to these cycles, through the collective actions of myriad
individual buy/sell decisions.  
%Similarly
In contrast, the genetic algorithms for which Holland is so famous
are exploratory models of \emph{mechanism}: they provide a simple
computational framework in which to explore the dynamics of Darwinian
evolution and whether the basic mechanisms of variation, differential
reproduction, and heredity are sufficient to account for the richness
of our natural world.

Holland was interested in models that explored basic principles and
mechanisms, even if they did not make specific or detailed
predictions.  Such models can show generically how certain behaviors
could be produced.  Holland pioneered a style of modeling that has
come to be known as ``individual-based'' or ``agent-based,'' in which
every component of a system is represented explicitly---e.g., every
trader in a stock market system or every cell in an immune system
model---and has a dynamic internal state. In such models, each agent has
its own behavior rules, which it can modify over time through learning.  In
order to capture the behavior of systems under spatial
constraints, these models are often defined over spatial structures,
such as networks or simple grids.

A given agent-based model encodes a theory about the mechanisms that
are relevant for producing the behavior of interest.  
%Similar to expert systems [MM says: ?]
%Such models are especially useful for studying systems
%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
%Agent-based models can facilitate interdisciplinary collaborations because the
%underlying rules can be easily communicated.  
[SF: THINKS this next text doesn't make sense until after complex
systems seciton.]
The agent-based models
championed by Holland were typically idealized versions of complex
systems and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

The emphasis on exploratory models to build intuitions was an
important theme of Holland's work, and he often quoted Eddington's
remark on the occasion of the first experimental test of Einstein's
theory of relativity: ``The contemplation in natural science of a wider
domain than the actual leads to a far better understanding of the
actual'' \cite{Eddington1927}.

It should be noted that Holland's view of modeling is by no means
typical.  For example, a textbook on computational modeling offers
 the following definition: ``Modeling is the application
of methods to analyze complex, real-world problems in order to make
predictions about what might happen with various actions''
\cite{Shiflet2014}.   This perspective rules out the
kind of exploratory modeling that Holland was most interested in.
